---
title: "Covid_19"
author: "Gerardo Goar, Oliver Tausendschön"
date: "2024-11-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Abstract



# Introduction


# Setup
```{r}
# Load dplyr for data manipulation
library(dplyr)
```



```{r}
covid_19<-read.csv("/home/oliver/Documents/Statisticcal_Modelling_Project/Data/covid_data.csv")
```


```{r}
head(covid_19)
```
Drop Columns that have information about Covid Included.

```{r}
covid_data_cleaned <- covid_19[, !(names(covid_19) %in% c("new_cases", "new_cases_smoothed", "total_deaths", "new_deaths", "new_deaths_smoothed", 
                                    "total_cases_per_million", "new_cases_per_million", "new_cases_smoothed_per_million", 
                                    "total_deaths_per_million", "new_deaths_per_million", "new_deaths_smoothed_per_million", 
                                    "reproduction_rate", "icu_patients", "icu_patients_per_million", "hosp_patients", 
                                    "hosp_patients_per_million", "weekly_icu_admissions", "weekly_icu_admissions_per_million", 
                                    "weekly_hosp_admissions", "weekly_hosp_admissions_per_million", "total_tests", "new_tests", 
                                    "total_tests_per_thousand", "new_tests_per_thousand", "new_tests_smoothed", 
                                    "new_tests_smoothed_per_thousand", "positive_rate", "tests_per_case", "tests_units", 
                                    "total_vaccinations", "people_vaccinated", "people_fully_vaccinated", "total_boosters", 
                                    "new_vaccinations", "new_vaccinations_smoothed", "total_vaccinations_per_hundred", 
                                    "people_vaccinated_per_hundred", "people_fully_vaccinated_per_hundred", 
                                    "total_boosters_per_hundred", "new_vaccinations_smoothed_per_million", 
                                    "new_people_vaccinated_smoothed", "new_people_vaccinated_smoothed_per_hundred"))]


covid_data_cleaned <- covid_data_cleaned %>%
  select(-starts_with("excess"))

```



```{r}

# Filter dataset and select the first date when total_cases is at least 1
covid_data_initial_cases <- covid_data_cleaned %>%
  filter(total_cases >= 1) %>%
  group_by(location) %>%
  arrange(date) %>%
  slice(1) %>%   # Keep only the first observation per country
  ungroup()

# View the cleaned dataset
head(covid_data_initial_cases)
```

```{r}
library(dplyr)

# Define the percentage threshold
threshold_pct <- 20  # Set percentage threshold, e.g., 20%
threshold <- ncol(covid_data_initial_cases) * threshold_pct / 100  # Calculate the number of NAs corresponding to the percentage

# Step 1: Remove rows with more missing values than the threshold
covid_data_initial_cases <- covid_data_initial_cases %>%
  filter(apply(., 1, function(x) sum(is.na(x)) <= threshold))  # Use apply to check each row

# Step 2: Impute remaining NAs with the median of each column
covid_data_initial_cases <- covid_data_initial_cases %>%
  mutate(across(everything(), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))

# View the cleaned and imputed data
print(covid_data_initial_cases)

```



```{r}

# Group by country, filter for the maximum total cases, and select the latest date if tied
covid_data_max_cases <- covid_data_cleaned %>%
  group_by(location) %>%
  filter(total_cases == max(total_cases, na.rm = TRUE)) %>%
  arrange(desc(date)) %>% 
  slice(1) %>% # Keep only the latest observation per country in case of ties
  ungroup()

# View the resulting dataset
head(covid_data_max_cases)

```

Merging these dataframes:

```{r}

covid_data_model<-merge(covid_data_initial_cases, covid_data_max_cases[, c("location", "total_cases")], by.x = "location", by.y = "location")

covid_data_model <- covid_data_model %>%
  select(-total_cases.x, -date)
#dummify continent:
library(dplyr)

# Create dummy variables manually
covid_data_model <- covid_data_model %>%
  mutate(
    continent_Africa = ifelse(continent == "Africa", 1, 0),
    continent_Asia = ifelse(continent == "Asia", 1, 0),
    continent_Europe = ifelse(continent == "Europe", 1, 0),
    continent_North_America = ifelse(continent == "North America", 1, 0),
    continent_Oceania = ifelse(continent == "Oceania", 1, 0),
    continent_South_America = ifelse(continent == "South America", 1, 0)
  ) %>%
  select(-continent)  # Optionally, remove the original 'continent' column

# View the resulting dataframe
head(covid_data_model)

```







```{r}
# Assuming your dataset is named df

# Step 1: Select columns 7 to 21 for interaction terms
df_selected <- covid_data_max_cases[, 7:21]  # Select columns 7 to 21

# Step 2: Create pairwise interactions
df_pairwise <- model.matrix(~ .^2, data = df_selected)  # This generates pairwise interactions

# Step 3: Create three-way interactions
#df_three_way <- model.matrix(~ .^3, data = df_selected)  # This generates pairwise + three-way interactions

# Step 4: Combine the interactions with the original dataset (if needed)
#df_with_interactions <- cbind(df, df_pairwise[, -1], df_three_way[, -(1:ncol(df_selected))])
df_with_interactions <- cbind(df, df_pairwise[, -1])  # Remove the intercept column
# The combined dataset now has the original features plus interactions
print(head(df_with_interactions))  # View the first few rows of the new dataset

```



We don't have enough data so we used the dataset combined with 'kaggle.com/datset'. We use the original dataset for the total number of cases and the new dataset to get the economic factors which we want to use for predicting. This effectively prevents us form having a lot of NaNs as in the original dataset.

```{r}
covid_19 <- read.csv("Data/World Data Set.csv")

```

```{r}
covid_19<-read.csv("Data/a524ed10-4f8e-4548-bdf2-ef21ef85df39_Data.csv")
```

```{r}
library(tidyverse)
df_wide <- covid_19 %>%
  pivot_wider(
    names_from = `Series.Name`,         # Column to use as new column names
    values_from = `X2019..YR2019.`,        # Column with values to fill those new columns
    id_cols = c(`Country.Name`, `Country.Code`)  # Columns to keep as identifiers
  )

# View the transformed data
print(df_wide)

```



# Data preperation



## Setup

