---
title: "Covid_19"
author: "Gerardo Goar, Oliver Tausendschön"
date: "2024-11-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Abstract


```{r}

```

# Introduction


# Setup
```{r}
# Load dplyr for data manipulation
library(dplyr)
```



```{r}
covid_19<-read.csv("/home/oliver/Documents/Statisticcal_Modelling_Project/Data/covid_data.csv")
```


```{r}
head(covid_19)
```
Drop Columns that have information about Covid Included.

```{r}
covid_data_cleaned <- covid_19[, !(names(covid_19) %in% c("new_cases", "new_cases_smoothed", "total_deaths", "new_deaths", "new_deaths_smoothed", 
                                    "total_cases_per_million", "new_cases_per_million", "new_cases_smoothed_per_million", 
                                    "total_deaths_per_million", "new_deaths_per_million", "new_deaths_smoothed_per_million", 
                                    "reproduction_rate", "icu_patients", "icu_patients_per_million", "hosp_patients", 
                                    "hosp_patients_per_million", "weekly_icu_admissions", "weekly_icu_admissions_per_million", 
                                    "weekly_hosp_admissions", "weekly_hosp_admissions_per_million", "total_tests", "new_tests", 
                                    "total_tests_per_thousand", "new_tests_per_thousand", "new_tests_smoothed", 
                                    "new_tests_smoothed_per_thousand", "positive_rate", "tests_per_case", "tests_units", 
                                    "total_vaccinations", "people_vaccinated", "people_fully_vaccinated", "total_boosters", 
                                    "new_vaccinations", "new_vaccinations_smoothed", "total_vaccinations_per_hundred", 
                                    "people_vaccinated_per_hundred", "people_fully_vaccinated_per_hundred", 
                                    "total_boosters_per_hundred", "new_vaccinations_smoothed_per_million", 
                                    "new_people_vaccinated_smoothed", "new_people_vaccinated_smoothed_per_hundred"))]

```

```{r}
covid_data_initial_cases <- covid_data_cleaned %>%
  filter(total_cases >= 1) %>%
  group_by(location) %>%
  arrange(date # Keep only the first observation per country
  ungroup()

# View the cleaned dataset
head(covid_data_initial_cases), descending = T) %>%
  slice(1) %>%  
```
```{r}

# Group by country, filter for the maximum total cases, and select the latest date if tied
covid_data_max_cases <- covid_data_cleaned %>%
  group_by(location) %>%
  filter(total_cases == max(total_cases, na.rm = TRUE)) %>%
  arrange(desc(date)) %>% 
  slice(1) %>% # Keep only the latest observation per country in case of ties
  ungroup()

# View the resulting dataset
head(covid_data_max_cases)

```
We don't have enough data so we used the dataset combined with 'kaggle.com'. We use the original dataset for the total number of cases and the new dataset to get the economic factors which we want to use for predicting. This effectively prevents us form having a lot of NaNs as in the original dataset.

```{r}
covid_19 <- read.csv("Data/covid_data.csv")

```
# Data preperation



## Setup

